---
title: "Getting Started with CompRisksBenchmaRk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CompRisksBenchmaRk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

**CompRisksBenchmaRk** provides a unified framework for benchmarking survival
models in the presence of competing risks. Its core features are:

- A **model registry** that lets you plug in any model with a common
  `fit(obj, args, ...)` / `predict_cif(fit_obj, newdata, time_grid)` interface.
- A **nested cross-validation** engine (`nested_cv_from_bench`) that reads the
  pre-built benchmark directory layout produced by the companion Python pipeline.
- A **metrics suite** covering Brier score, IBS, AUC, C-index, calibration
  plots, O/E ratio, and decision curve analysis.
- Built-in support for **multiple imputation** (median/mode or MICE) with
  mean-pooling of CIF predictions across imputed datasets.

Built-in models: `"FGR"`, `"FGRP"`, `"csCPH"`, `"RSF"`.

---

## Installation

```r
# From a local source directory:
devtools::install_local("path/to/CompRisksBenchmaRk")
```

---

## 1  Example data: `survival::pbc`

We use the Mayo Clinic Primary Biliary Cholangitis (PBC) trial from the
**survival** package. The endpoint has two competing outcomes:

- `status = 1` — liver transplant (cause 1)
- `status = 2` — death (cause 2)
- `status = 0` — censored

Time is recorded in days; we convert to years for readability.

```{r data}
library(CompRisksBenchmaRk)
library(survival)

data(pbc, package = "survival")
pbc2        <- pbc[!is.na(pbc$trt), ]         # randomised patients only
pbc2$time   <- pbc2$time / 365.25             # days -> years
pbc2$stage  <- as.factor(                     # group stages 1-2 vs 3 vs 4
  ifelse(pbc2$stage %in% c(1, 2), "1-2", as.character(pbc2$stage))
)

feature_cols <- c("age", "stage", "bili", "albumin", "protime")
pbc2         <- pbc2[, c("time", "status", feature_cols)]
pbc2         <- pbc2[complete.cases(pbc2), ]

table(pbc2$status)
table(pbc2$stage)
```

`cr_data()` validates inputs, coerces the event column, records covariate
types, and sorts by time:

```{r prepare}
cr <- cr_data(pbc2, time_var = "time", event_var = "status")
cr
```

---

## 2  The model registry

```{r registry}
list_cr_models()
get_cr_model("FGR")$info()
```

### Registering a custom model

Any model whose `fit` accepts a `cr_data` object and whose
`predict_cif` returns a `[n, K, Tm]` array can be registered.

```{r custom-model, eval = FALSE}
register_cr_model(
  key = "MyModel",
  fit = function(obj, args = list(), ...) {
    # ... train model using obj@data, obj@time_var, obj@event_var, obj@causes ...
    list(causes = obj@causes, fit = my_fit, model_key = "MyModel")
  },
  predict_cif = function(fit_obj, newdata, time_grid) {
    n  <- nrow(newdata)
    K  <- length(fit_obj$causes)
    Tm <- length(time_grid)
    array(0.1, dim = c(n, K, Tm))
  },
  info = function() list(
    name         = "My Custom Model",
    supports     = "CIF",
    needs_tuning = FALSE,
    default_grid = function() tibble::tibble()
  )
)
```

---

## 3  Fit and predict manually

Models are fitted by passing a `cr_data` object.  For demonstration
purposes we fit and predict on the full dataset; `predict_cif()` dispatches
internally via `fit_obj$model_key` and returns an `[n, K, Tm]` array.

```{r fit-predict, warning = FALSE}
# Cap time grid at the observed data range to avoid extrapolation
max_time  <- floor(max(cr@data[[cr@time_var]]))
time_grid <- seq(1, max_time, by = 1)

# --- FGR ---
fit_FGR  <- fit_cr_model("FGR", cr)
cif_FGR  <- predict_cif(fit_FGR,  newdata = cr, time_grid = time_grid)

# --- csCPH ---
# Note: riskRegression::predictRisk may warn about predictions outside [0,1]
# when predicting on training data with extreme covariate combinations.
fit_csCPH <- fit_cr_model("csCPH", cr)
cif_csCPH <- predict_cif(fit_csCPH, newdata = cr, time_grid = time_grid)

# --- RSF ---
fit_RSF  <- fit_cr_model("RSF", cr)
cif_RSF  <- predict_cif(fit_RSF,  newdata = cr, time_grid = time_grid)

# --- FGRP (requires a regularisation parameter) ---
fit_FGRP <- fit_cr_model("FGRP", cr, args = list(lambda_seq = 0.01))
cif_FGRP <- predict_cif(fit_FGRP, newdata = cr, time_grid = time_grid)

# Collect for downstream use; all arrays: n x 2 causes x 10 time points
model_keys <- c("FGR", "csCPH", "RSF", "FGRP")
fits <- list(FGR = fit_FGR, csCPH = fit_csCPH, RSF = fit_RSF, FGRP = fit_FGRP)
cifs <- list(FGR = cif_FGR, csCPH = cif_csCPH, RSF = cif_RSF, FGRP = cif_FGRP)

sapply(cifs, dim)
```

The explicit calls above make it straightforward to inspect or debug any
individual model.  For convenience, the same result can be obtained
programmatically:

```{r fit-predict-lapply, eval = FALSE}
fits <- lapply(setNames(model_keys, model_keys), function(key) {
  args <- if (key == "FGRP") list(lambda_seq = 0.01) else list()
  fit_cr_model(key, obj = cr, args = args)
})
cifs <- lapply(fits, function(fit)
  predict_cif(fit, newdata = cr, time_grid = time_grid)
)
```

---

## 4  Scoring

```{r scoring}
# --- FGR ---
scores_brier_FGR <- calculate_metrics(
  cr         = cr,
  eval_times = time_grid,
  cif        = cif_FGR,
  metrics = "Brier",
  summary = "IBS"
)
scores_auc_FGR <- calculate_metrics(
  cr         = cr,
  eval_times = time_grid,
  cif        = cif_FGR,
  metrics = "tdAUC"
)

# --- csCPH ---
scores_brier_csCPH <- calculate_metrics(
  cr         = cr,
  eval_times = time_grid,
  cif        = cif_csCPH,
  metrics = "Brier",
  summary = "IBS"
)
scores_auc_csCPH <- calculate_metrics(
  cr         = cr,
  eval_times = time_grid,
  cif        = cif_csCPH,
  metrics = "tdAUC"
)

# --- RSF ---
scores_brier_RSF <- calculate_metrics(
  cr         = cr,
  eval_times = time_grid,
  cif        = cif_RSF,
  metrics = "Brier",
  summary = "IBS"
)
scores_auc_RSF <- calculate_metrics(
  cr         = cr,
  eval_times = time_grid,
  cif        = cif_RSF,
  metrics = "tdAUC"
)

# --- FGRP ---
scores_brier_FGRP <- calculate_metrics(
  cr         = cr,
  eval_times = time_grid,
  cif        = cif_FGRP,
  metrics = "Brier",
  summary = "IBS"
)
scores_auc_FGRP <- calculate_metrics(
  cr         = cr,
  eval_times = time_grid,
  cif        = cif_FGRP,
  metrics = "tdAUC"
)

# IBS per cause for each model
scores_brier_FGR$IBS
scores_brier_FGR$Brier$cause_2    # time-specific Brier score for death (FGR)
```

For convenience, the same scoring can be obtained programmatically:

```{r scoring-lapply, eval = FALSE}
scores <- lapply(cifs, function(cif)
  calculate_metrics(
    cr      = cr,
    eval_times = time_grid,
    cif     = cif,
      metrics = c("Brier", "tdAUC"),
    summary = "IBS"
  )
)

# Compare IBS for cause 2 across models
sapply(scores, function(s) s$IBS$cause_2)
```

---

## 5  Calibration

```{r calibration, fig.alt="Calibration plots for cause 2 (death) at 5 years"}
horizon <- 5
hi      <- which.min(abs(time_grid - horizon))

# Calibration plot for each model at the 5-year horizon for cause 2 (death)
cal_list <- lapply(model_keys, function(key) {
  CalibrationPlot(
    model_name       = key,
    predictions      = cifs[[key]][, 2, hi],
    data             = cr@data,
    time             = cr@data[[cr@time_var]],
    status           = cr@data[[cr@event_var]],
    tau              = horizon,
    cause            = 2,
    predictions.type = "CIF",
    loess_smoothing  = TRUE,
    graph            = TRUE
  )
})
names(cal_list) <- model_keys

lapply(cal_list, function(cal) print(cal$graphs[[1]]))
cal_list$FGR$calib.measures[[1]]    # ICI, E50, E90, Emax, RSB for FGR
cal_list$FGR$OE_summary             # O/E ratio with 95% CI for FGR
```

---

## 6  Decision curve analysis

```{r dca, fig.alt="Decision curve analysis for 5-year death risk"}
# Add each model's 5-year cause-2 predictions as columns for DCA
test_dca <- cr@data
for (key in model_keys)
  test_dca[[paste0("pred_", key)]] <- cifs[[key]][, 2, hi]

pred_cols <- paste0("pred_", model_keys)

du <- ClinicalUtility(
  data         = test_dca,
  outcome      = cr@event_var,
  ttoutcome    = cr@time_var,
  timepoint    = horizon,
  predictors   = pred_cols,
  graph        = TRUE,
  intervention = FALSE,
  cmprsk       = TRUE
)

print(du$plot)
```

---

## 7  Nested cross-validation (benchmark workflow)

`nested_cv_from_bench()` reads the pre-built directory layout produced by the
companion Python pipeline. It constructs a `cr_data` object internally for each fold
(using the standardised `"time"` / `"event"` column names) before calling
each model's `fit()` function.

```{r nested-cv, eval = FALSE}
results <- nested_cv_from_bench(
  out_dir   = "path/to/BenchResults/pbc",
  model_key = "FGR",
  seed      = 123,
  verbose   = TRUE
)

# RSF with hyperparameter tuning
results_rsf <- nested_cv_from_bench(
  out_dir   = "path/to/BenchResults/pbc",
  model_key = "RSF",
  grid      = list(ntree = c(200, 500), nodesize = c(10, 20)),
  seed      = 123,
  verbose   = TRUE
)
```

After all models have been run, collect and score with `get_results()`:

```{r get-results, eval = FALSE}
out <- get_results(
  data_root         = "path/to/BenchResults",
  dataset           = "pbc",
  r_models          = c("FGR", "csCPH", "RSF"),
  python_models     = c("DeepHit", "DeSurv"),
  outer_folds       = 5,
  horizon           = 5.0,
  seed              = 42,
  cause_of_interest = 2,
  sample_curves     = FALSE
)

lapply(out$res$FGR, function(f) f$IBS$cause_2)
```

---

## 8  Imputation

`impute_split()` learns all imputation parameters from the training set to
prevent data leakage. For MICE (`m > 1`), predictions are pooled across
imputed datasets with `pool_cifs_mean()`.

```{r imputation, eval = FALSE}
feature_cols <- cr@covars$covars_names

imp <- impute_split(
  train        = cr@data,
  test         = cr@data,
  feature_cols = feature_cols,
  method       = "median_mode"    # or "mice" for multiple imputation
)

cif_list <- lapply(seq_len(imp$m), function(k) {
  tr_k <- imp$imputations[[k]]$train
  te_k <- imp$imputations[[k]]$test
  cr_k <- cr_data(tr_k, time_var = cr@time_var,
                  event_var = cr@event_var)
  fit_k <- fit_cr_model("FGR", cr_k)
  cr_te_k <- cr_data(te_k, time_var = cr@time_var, event_var = cr@event_var)
  predict_cif(fit_k, newdata = cr_te_k, time_grid = time_grid)
})

cif_pooled <- pool_cifs_mean(cif_list)
```

---

## Session info

```{r session-info}
sessionInfo()
```
