---
title: "Getting Started with CompRisksBenchmaRk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CompRisksBenchmaRk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

**CompRisksBenchmaRk** provides a unified framework for benchmarking survival
models in the presence of competing risks.  Its core features are:

- A **model registry** that lets you plug in any model with a common
  `fit` / `predict_cif` interface.
- A **nested cross-validation** engine (`nested_cv_from_bench`) that reads the
  pre-built benchmark directory layout produced by the companion Python pipeline.
- A **metrics suite** covering Brier score, IBS, AUC, C-index, calibration
  plots, O/E ratio, and decision curve analysis.
- Built-in support for **multiple imputation** (median/mode or MICE) with
  Rubin's-rules pooling of CIF predictions.

Built-in models: `"FGR"`, `"FGRP"`, `"csCPH"`, `"RSF"`, `"VM_spline"`,
`"VM_piecewise"`.

---

## Installation

```r
# From a local zip / source directory:
install.packages("path/to/CompRisksBenchmaRk_0.1.0.tar.gz",
                 repos = NULL, type = "source")

# Or with devtools:
devtools::install_local("path/to/CompRisksBenchmaRk")
```

---

## 1  Simulate data

`sim_cmprks()` generates a small competing risks dataset useful for
exploration.

```{r simulate}
library(CompRisksBenchmaRk)

set.seed(42)
dat <- sim_cmprks(n = 400, seed = 42)

head(dat)
table(dat$event)   # 0 = censored, 1 = cause 1, 2 = cause 2
```

---

## 2  The model registry

All models are stored in a package-internal environment and accessed by a
character key.

```{r registry}
# See what is registered after loading the package
list_cr_models()

# Inspect a model's metadata
mdl  <- get_cr_model("FGR")
mdl$info()
```

### Registering a custom model

Any model that can produce a CIF array `[n, K, Tm]` can be registered:

```{r custom-model, eval = FALSE}
register_cr_model(
  key = "MyModel",

  fit = function(x, y_time, y_event, args = list()) {
    # ... train your model ...
    structure(list(fit = my_fit, causes = causes),
              class = c("cr_model_mymodel", "cr_model"))
  },

  predict_cif = function(fit_obj, x_new, times) {
    # Return array dim c(nrow(x_new), K, length(times))
  },

  info = function() list(
    name         = "My Custom Model",
    supports     = "CIF",
    needs_tuning = FALSE,
    default_grid = function() tibble::tibble()
  )
)
```

---

## 3  Fit and predict manually

```{r fit-predict}
library(survival)

# Prepare data
dat <- prepare_data(dat, time_col = "time", event_col = "event")

causes <- sort(unique(dat$event[dat$event != 0]))
times  <- quantile(dat$time[dat$event != 0], probs = seq(0.1, 0.9, by = 0.1))
times  <- sort(unique(round(times, 2)))

# Train / test split
set.seed(1)
idx_tr  <- sample(nrow(dat), 300)
train   <- dat[ idx_tr, ]
test    <- dat[-idx_tr, ]

feature_cols <- setdiff(names(dat), c("time", "event"))

# Build model matrices
dm     <- remake_X(train, test, feature_cols)
x_tr   <- dm$x_train
x_te   <- dm$x_new

# Fit FGR
mdl    <- get_cr_model("FGR")
fit    <- mdl$fit(x = x_tr, y_time = train$time,
                  y_event = train$event, args = list())

# Predict CIFs  →  array [n_test, K, Tm]
cif    <- mdl$predict_cif(fit, x_new = x_te, times = times)

dim(cif)   # n_test × 2 causes × n_times
```

---

## 4  Scoring

`score_from_cifs()` wraps `riskRegression::Score`, `pec::cindex`, and the
package's own calibration functions into a single call.

```{r scoring}
scores <- score_from_cifs(
  out        = cif,
  test       = test,
  times      = times,
  causes     = causes,
  metrics    = c("brier", "auc"),
  summary    = "ibs"
)

# IBS per cause
scores$cause_ibs
```

### Brier score at individual time points

```{r brier}
# Per-time-point Brier score for cause 1
scores$cause_bs$cause_1
```

---

## 5  Calibration

`CalibrationPlot()` uses jackknife pseudovalues (Aalen-Johansen) smoothed by
LOESS to assess calibration at a chosen horizon.

```{r calibration, fig.alt="Calibration plot for cause 1"}
horizon <- median(times)

# Predictions at the horizon index
hi      <- which.min(abs(times - horizon))
pred_h  <- cif[, 1, hi]   # cause 1 predictions for all test subjects

cal <- CalibrationPlot(
  model_name       = "FGR",
  predictions      = pred_h,
  data             = test,
  time             = test$time,
  status           = test$event,
  tau              = horizon,
  cause            = 1,
  predictions.type = "CIF",
  loess_smoothing  = TRUE,
  graph            = TRUE
)

print(cal$graphs[[1]])
cal$calib.measures[[1]]   # ICI, E50, E90, Emax, RSB
cal$OE_summary            # observed / expected ratio
```

---

## 6  Decision curve analysis

```{r dca, fig.alt="Decision curve analysis"}
# Add predictions as a column for ClinicalUtility
test_dca <- test
test_dca$pred_FGR <- pred_h

du <- ClinicalUtility(
  data        = test_dca,
  outcome     = "event",
  ttoutcome   = "time",
  timepoint   = horizon,
  predictors  = "pred_FGR",
  graph       = TRUE,
  intervention = FALSE,
  cmprsk      = TRUE
)

print(du$plot)
```

---

## 7  Nested cross-validation (benchmark workflow)

When the benchmark directory layout has been produced by the companion Python
pipeline, use `nested_cv_from_bench()`:

```{r nested-cv, eval = FALSE}
# Run FGR on a pre-built benchmark directory
results <- nested_cv_from_bench(
  out_dir   = "path/to/BenchResults/my_dataset",
  model_key = "FGR",
  seed      = 123,
  verbose   = TRUE
)

# Run RSF with a hyperparameter grid
results_rsf <- nested_cv_from_bench(
  out_dir   = "path/to/BenchResults/my_dataset",
  model_key = "RSF",
  grid      = list(
    ntree    = c(200L, 500L),
    nodesize = c(10L, 20L)
  ),
  seed    = 123,
  verbose = TRUE
)
```

After all models have been run, collect and score predictions with
`get_results()`:

```{r get-results, eval = FALSE}
out <- get_results(
  data_root        = "path/to/BenchResults",
  dataset          = "my_dataset",
  r_models         = c("FGR", "csCPH", "RSF"),
  python_models    = c("DeepHit", "DeSurv"),
  outer_folds      = 5,
  horizon          = 3.0,
  seed             = 42,
  cause_of_interest = 1L,
  sample_curves    = FALSE
)

# Per-fold IBS for FGR, cause 1
lapply(out$res$FGR, function(f) f$cause_ibs$cause_1)
```

---

## 8  Imputation

`impute_split()` handles missing data before model fitting.  The method is
learnt from the training set only, preventing data leakage.

```{r imputation, eval = FALSE}
imp <- impute_split(
  train        = train,
  test         = test,
  feature_cols = feature_cols,
  method       = "median_mode"   # or "mice" for multiple imputation
)

# For MICE (m > 1), pool predictions across imputed datasets:
cif_list <- lapply(seq_len(imp$m), function(k) {
  dm_k <- remake_X(imp$imputations[[k]]$train,
                   imp$imputations[[k]]$test,
                   feature_cols)
  fit_k <- mdl$fit(x = dm_k$x_train,
                   y_time  = imp$imputations[[k]]$train$time,
                   y_event = imp$imputations[[k]]$train$event)
  mdl$predict_cif(fit_k, dm_k$x_new, times)
})

cif_pooled <- pool_cifs_mean(cif_list)
```

---

## Session info

```{r session-info}
sessionInfo()
```
