---
title: "Getting Started with CompRisksBenchmaRk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CompRisksBenchmaRk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

**CompRisksBenchmaRk** provides a unified framework for benchmarking survival
models in the presence of competing risks. Its core features are:

- A **model registry** that lets you plug in any model with a common
  `fit(cr, args, ...)` / `predict_cif(fit_obj, newdata, time_grid)` interface.
- A **nested cross-validation** engine (`nested_cv_from_bench`) that reads the
  pre-built benchmark directory layout produced by the companion Python pipeline.
- A **metrics suite** covering Brier score, IBS, AUC, C-index, calibration
  plots, O/E ratio, and decision curve analysis.
- Built-in support for **multiple imputation** (median/mode or MICE) with
  mean-pooling of CIF predictions across imputed datasets.

Built-in models: `"FGR"`, `"FGRP"`, `"csCPH"`, `"RSF"`.

---

## Installation

```r
# From a local source directory:
devtools::install_local("path/to/CompRisksBenchmaRk")
```

---

## 1  Example data: `survival::pbc`

We use the Mayo Clinic Primary Biliary Cholangitis (PBC) trial from the
**survival** package. The endpoint has two competing outcomes:

- `status = 1` — liver transplant (cause 1)
- `status = 2` — death (cause 2)
- `status = 0` — censored

Time is recorded in days; we convert to years for readability.

```{r data}
library(CompRisksBenchmaRk)
library(survival)

data(pbc, package = "survival")
pbc2        <- pbc[!is.na(pbc$trt), ]         # randomised patients only
pbc2$time   <- pbc2$time / 365.25             # days -> years

feature_cols <- c("age", "bili", "albumin", "protime", "platelet")
pbc2         <- pbc2[, c("time", "status", feature_cols)]
pbc2         <- pbc2[complete.cases(pbc2), ]

table(pbc2$status)
```

`cr_data()` validates inputs, coerces the event column, records covariate
types, and sorts by time:

```{r prepare}
cr <- cr_data(pbc2, time_var = "time", event_var = "status")
head(cr@data[, c("time", "status")])
cr@causes
cr@covars
```

---

## 2  The model registry

```{r registry}
list_cr_models()
get_cr_model("FGR")$info()
```

### Registering a custom model

Any model whose `fit` accepts a `cr_data` object and whose
`predict_cif` returns a `[n, K, Tm]` array can be registered.

```{r custom-model, eval = FALSE}
register_cr_model(
  key = "MyModel",
  fit = function(cr, args = list(), ...) {
    # ... train model using cr@data, cr@time_var, cr@event_var, cr@causes ...
    list(causes = cr@causes, fit = my_fit, model_key = "MyModel")
  },
  predict_cif = function(fit_obj, newdata, time_grid) {
    n  <- nrow(newdata)
    K  <- length(fit_obj$causes)
    Tm <- length(time_grid)
    array(0.1, dim = c(n, K, Tm))
  },
  info = function() list(
    name         = "My Custom Model",
    supports     = "CIF",
    needs_tuning = FALSE,
    default_grid = function() tibble::tibble()
  )
)
```

---

## 3  Fit and predict manually

Models are fitted by passing a `cr_data` object.

```{r fit-predict}
time_grid <- seq(1, 10, by = 1)   # evaluation times: years 1-10

set.seed(1)
idx_tr   <- sample(nrow(cr@data), floor(0.75 * nrow(cr@data)))
cr_train <- cr[idx_tr]
cr_test  <- cr[-idx_tr]

mdl <- get_cr_model("FGR")
fit <- mdl$fit(cr = cr_train, args = list())

# predict_cif() wrapper coerces newdata and time_grid before dispatching
cif <- predict_cif(fit, newdata = cr_test@data, time_grid = time_grid)
dim(cif)   # n_test x 2 causes x 10 time points
```

---

## 4  Scoring

```{r scoring}
causes <- sort(unique(dat$event[dat$event != 0]))   # 1, 2

scores <- score_from_cifs(
  out     = cif,
  test    = test,
  times   = time_grid,
  causes  = causes,
  metrics = c("brier", "auc"),
  summary = "ibs"
)

scores$cause_ibs           # IBS per cause
scores$cause_bs$cause_2    # time-specific Brier score for death
```

---

## 5  Calibration

```{r calibration, fig.alt="Calibration plot for cause 2 (death) at 5 years"}
horizon   <- 5
hi        <- which.min(abs(time_grid - horizon))
pred_dead <- cif[, 2, hi]

cal <- CalibrationPlot(
  model_name       = "FGR",
  predictions      = pred_dead,
  data             = test,
  time             = test$time,
  status           = test$event,
  tau              = horizon,
  cause            = 2,
  predictions.type = "CIF",
  loess_smoothing  = TRUE,
  graph            = TRUE
)

print(cal$graphs[[1]])
cal$calib.measures[[1]]    # ICI, E50, E90, Emax, RSB
cal$OE_summary             # O/E ratio with 95% CI
```

---

## 6  Decision curve analysis

```{r dca, fig.alt="Decision curve analysis for 5-year death risk"}
test_dca          <- test
test_dca$pred_FGR <- pred_dead

du <- ClinicalUtility(
  data         = test_dca,
  outcome      = "event",
  ttoutcome    = "time",
  timepoint    = horizon,
  predictors   = "pred_FGR",
  graph        = TRUE,
  intervention = FALSE,
  cmprsk       = TRUE
)

print(du$plot)
```

---

## 7  Nested cross-validation (benchmark workflow)

`nested_cv_from_bench()` reads the pre-built directory layout produced by the
companion Python pipeline. It constructs a `cr_data` object internally for each fold
(using the standardised `"time"` / `"event"` column names) before calling
each model's `fit()` function.

```{r nested-cv, eval = FALSE}
results <- nested_cv_from_bench(
  out_dir   = "path/to/BenchResults/pbc",
  model_key = "FGR",
  seed      = 123,
  verbose   = TRUE
)

# RSF with hyperparameter tuning
results_rsf <- nested_cv_from_bench(
  out_dir   = "path/to/BenchResults/pbc",
  model_key = "RSF",
  grid      = list(ntree = c(200, 500), nodesize = c(10, 20)),
  seed      = 123,
  verbose   = TRUE
)
```

After all models have been run, collect and score with `get_results()`:

```{r get-results, eval = FALSE}
out <- get_results(
  data_root         = "path/to/BenchResults",
  dataset           = "pbc",
  r_models          = c("FGR", "csCPH", "RSF"),
  python_models     = c("DeepHit", "DeSurv"),
  outer_folds       = 5,
  horizon           = 5.0,
  seed              = 42,
  cause_of_interest = 2,
  sample_curves     = FALSE
)

lapply(out$res$FGR, function(f) f$cause_ibs$cause_2)
```

---

## 8  Imputation

`impute_split()` learns all imputation parameters from the training set to
prevent data leakage. For MICE (`m > 1`), predictions are pooled across
imputed datasets with `pool_cifs_mean()`.

```{r imputation, eval = FALSE}
imp <- impute_split(
  train        = train,
  test         = test,
  feature_cols = feature_cols,
  method       = "median_mode"    # or "mice" for multiple imputation
)

cif_list <- lapply(seq_len(imp$m), function(k) {
  tr_k <- imp$imputations[[k]]$train
  te_k <- imp$imputations[[k]]$test
  cr_k  <- cr_data(tr_k, time_var = "time", event_var = "event")
  fit_k <- mdl$fit(cr = cr_k,
  mdl$predict_cif(fit_k, newdata = te_k, time_grid = time_grid)
})

cif_pooled <- pool_cifs_mean(cif_list)
```

---

## Session info

```{r session-info}
sessionInfo()
```
